---
title: "Expected Goals Model for 2023 WWC"
author: "Hope Donoghue"
format: gfm
---

Which area on the soccer field has the highest probability of a goal occurring? This project explores this question along with many others related to goal probabilities in soccer. For this project, I built an expected goals model for the 2023 Women's World Cup using logistic regression, and created a variety of visuals like the one shown below. An expected goals model predicts the probability that a shot will result in a goal based on the characteristics of that shot and the events leading up to it. The data that I used for this project comes from StatsBomb and it is event data. In addition to building the model, I also built a Shiny App that displays the expected goal probabilities (xG values) for each shot that a player took in the 2023 Women's World Cup. Overall, I found that higher probability shots come from within the 18 yard box and that xG is an effective metric for measuring a player's performance.   

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# read in libraries
library(readr)
library(dplyr)
library(knitr)
library(tidyverse)
library(plotly)
library(ggsoccer)
library(broom)
library(stringr)
library(modelr)
library(SBpitch)
library(ggplot2)
library(caret)
```


```{r, echo = FALSE, message = FALSE, warning = FALSE}
# read in the csv file that contains the cleaned up data set for all shots in WWC 2023
shots_valid_wwc2023_new <- read_csv("shots_valid_wwc2023_new.csv")
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
shots_valid_wwc_log <- read_csv("shots_valid_wwc_log.csv")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# using an outline of a soccer pitch 
pitch = create_Pitch("#ffffff", "#A9A9A9", "#ffffff", "#000000", BasicFeatures = FALSE, goaltype = "box")
```

```{r, echo = FALSE, warning= FALSE, message= FALSE}
# showing the shot locations of all shots in WWC 2023 to answer
# where the most probable location of scoring a goal is
# using open point shape to include stroke and color to the shot map

shots_valid_wwc_log_new <- shots_valid_wwc_log %>%
  mutate(is.goal_numeric = if_else(is.goal == "Goal", 1.5, 0.5))

a8 = pitch + geom_point(data = shots_valid_wwc_log_new, aes(x = location.x, y = location.y, color = .fitted, stroke = is.goal_numeric), shape = 1, size = 2) + labs(color = "Probability of Goal \n (xG)", caption = "Note: Removed Republic of Ireland's goal directly from corner \n Note: Darker filled in circles mean goal was scored") + scale_colour_gradientn(colors = c(low = "blue", high = "red"), limits=c(0, 1)) + theme(plot.title = element_text(face = "bold", size = 20))

a8
```

